---
title: "Reproducible Research - Week 2 Assignment"
author: "Caio Rodrigues"
date: "September 7th, 2020"
output: html_document
---

### Loading and Preprocessing the data

```{r loading data, echo = TRUE, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)

if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = "./data/activitydata.zip", method = "curl")
unzip(zipfile = "./data/activitydata.zip", exdir = "./data")
activity <- read.csv("./data/activity.csv")
```

### What is the mean total number of steps taken per day?

1. Making a histogram of the total number of steps taken each day
```{r steps per day, echo = TRUE, message = FALSE, warning = FALSE}
activity$date <- as.Date(activity$date)
stepsperday <- group_by(activity, date) %>%
     summarize(total_steps = sum(steps, na.rm = TRUE))

qplot(total_steps, data = stepsperday, na.rm = TRUE, 
      binwidth = 500, 
      xlab = "Total number of steps per day",
      ylab = "Frequency",
      main = "Histogram of the total number of steps taken each day")
```

2. Calculate the **mean** and **median** total number of steps taken per day
```{r mean steps per day, echo = TRUE, message = FALSE, warning = FALSE}
meansteps <- stepsperday %>%
     summarize(average = mean(total_steps, na.rm = TRUE), median = median(total_steps, na.rm = TRUE))
meansteps

```

### What is the average daily activity pattern?
1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r daily activity pattern, echo = TRUE, message = FALSE, warning = FALSE}
dailyactivity <- activity %>% group_by(interval) %>% 
     summarize(average = mean(steps,na.rm = TRUE))

qplot(interval, average,
      data = dailyactivity,geom="line",
      xlab = "5-minute intervals",
      ylab = "Average steps taken across all days")

```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r max steps interval, echo = TRUE, message = FALSE, warning = FALSE}
dailyactivity[which.max(dailyactivity$average),]

```

### Imputing missing values
1. Calculate the total number of missing values
```{r NA number, echo = TRUE, message = FALSE, warning = FALSE}
activityNA <- which(is.na(activity$steps))
length(activityNA)

```

2. Devise a strategy for filling in all of the missing values in the dataset
```{r filling NA, echo = TRUE, message = FALSE, warning = FALSE}
activity_noNA <- activity[which(!is.na(activity$steps)),]
activity_avg <- activity_noNA %>%
     group_by(interval) %>%
     summarize(average = mean(steps))

activity_avg$average <- as.integer(activity_avg$average)

activityNA <- activity[which(is.na(activity$steps)),]

activityNA$steps <- ifelse(activityNA$interval == activity_avg$interval, activity_avg$average)

```

3. Create a new dataset with missing values filled in
```{r non-NA dataset, echo = TRUE, message = FALSE, warning = FALSE}
activity_NA_filled <- rbind(activity_noNA, activityNA) %>% arrange(date)
```

4. Make new histogram with imputed values and report **mean** and **median** total number of steps taken per day
```{r non-NA histogram, echo = TRUE, message = FALSE, warning = FALSE}
stepsperday2 <- group_by(activity_NA_filled, date) %>%
     summarize(total_steps = sum(steps))

qplot(total_steps, data = stepsperday2, 
      binwidth = 500, 
      xlab = "Total number of steps per day",
      ylab = "Frequency",
      main = "Histogram of the total number of steps taken each day (missing values imputed)")

meansteps2 <- stepsperday2 %>%
     summarize(average = mean(total_steps, na.rm = TRUE), median = median(total_steps, na.rm = TRUE))
meansteps2
```

###Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r weekday variable, echo = TRUE, message = FALSE, warning = FALSE}
activity$date <- as.Date(strptime(activity$date, format="%Y-%m-%d"))

activity$datetype <- sapply(activity$date, function(x) {
     if (weekdays(x) == "Saturday" | weekdays(x) =="Sunday") 
     {y <- "Weekend"} else 
     {y <- "Weekday"}
     y
})
```

2. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r weekday plot, echo = TRUE, message = FALSE, warning = FALSE}
activity_weekday <- aggregate(steps~interval + datetype, activity, mean, na.rm = TRUE)
plot<- ggplot(activity_weekday, aes(x = interval , y = steps, color = datetype)) +
       geom_line() +
       labs(title = "Average daily steps by type of date", x = "Interval", y = "Average number of steps") +
       facet_wrap(~datetype, ncol = 1, nrow=2)
print(plot)

```
